---
title: "Tidyverse basics with DatasauRus"
author: "Elisa Gomez de Lope"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

This guided practical will demonstrate that the **tidyverse** allows to compute summary statistics and visualize datasets efficiently.
This dataset is already stored in a **tidy** `tibble`, cleaning steps will come in future practicals.


##### Those kind of questions are optional {.bonus}

## `datasauRus` package

- check if you have the package `datasauRus` and `tidyverse` installed

```{r}
library(datasauRus)
library(tidyverse)
```

- should return nothing. If `there is no package called ‘datasauRus’` appears, it means that the package needs 
to be installed. Use this:
```{r}
# install.packages("datasauRus")
```


## Explore the dataset

Since we are dealing with a `tibble`, we can just type
```{r}
datasaurus_dozen
```

only the first **10** rows are displayed.


##### What are the dimensions of this dataset? Rows and columns?

- **base** version, using either `dim()`, `ncol()` and `nrow()`

```{r}
dim(datasaurus_dozen)
ncol(datasaurus_dozen)
nrow(datasaurus_dozen)
```


- **tidyverse** version

```{r}
tibble(datasaurus_dozen)
```


##### Assign the `datasaurus_dozen` to the `ds_dozen` name This aims at populating the **Global Environment**

```{r}
ds_dozen <- datasaurus_dozen
```


##### Using Rstudio, those dimensions are now also reported within the interface, where?

In the global environment


## How many datasets are present?

- **base** version

```{block, opts.label = "tip"}
you want to count the number of **unique** elements in the column **dataset**.
The function **length()** returns the length of a vector, such as the unique elements
```

```{r}
unique(ds_dozen$dataset) %>% length()
```


- **tidyverse** version

```{r}
# n_distinct counts the unique elements in a given vector.
# we use summarise to return only the desired column named n here.
summarise(ds_dozen, n = n_distinct(dataset))
```

- even better way, compute and display the number of lines per `dataset`

```{block, opts.label = "tip"}
the function `count` in `dplyr` does the `group_by()` by the specified column + `summarise(n = n())` which returns the number of observation per defined group.
```

```{r}
count(ds_dozen, dataset)

```



## Check summary statistics per dataset

##### Compute the mean of the `x` & `y` column. For this, you need to `group_by()` the appropriate column and then `summarise()`

```{block, opts.label = "tip"}
in `summarise()` you can define as many new columns as you wish. No need to call it for every single variable.
```

```{r message=TRUE, warning=TRUE, include=FALSE, paged.print=FALSE}
ds_dozen %>%
  group_by(dataset) %>%
  summarise(mean_x = mean(x),
            mean_y = mean(y))
```


```{r message=TRUE, warning=TRUE, include=FALSE, paged.print=FALSE}
ds_dozen %>%
  group_by(dataset) %>%
  summarise(sd_x = sd(x),
            sd_y = sd(y))
```



##### Compute both mean and standard deviation (sd) in one go using `across()` {.bonus}


```{r}
ds_dozen %>%
  group_by(dataset) %>%
  # across works with first on which columns and second on what to perform on selection
  # 2 possibilities to select columns
  # summarise(across(where(is.double), list(mean = mean, sd = sd)))
  summarise(across(c(x, y), list(mean = mean, sd = sd)))
```


##### What can you conclude?

They look all similar based on summary stats. The mean and sd are the same in all datasets.

## Plot the _datasauRus_

##### Plot the `ds_dozen` with `ggplot` such the **aesthetics** are `aes(x = x, y = y)` 
with the **geometry** `geom_point()`

```{block, opts.label = "tip"}
the `ggplot()` and `geom_point()` functions must be linked with a **+** sign
```

```{r}
ds_dozen %>%
  ggplot(aes(x=x, y =y)) +
  geom_point()
```


##### Reuse the above command, and now colored by the `dataset` column

```{r}
ds_dozen %>%
  ggplot(aes(x=x, y =y, colour=dataset)) +
  geom_point()

```


Too many datasets are displayed.

##### How can we plot only one at a time?

```{block, opts.label = "tip"}
You can filter for one dataset upstream of plotting
```

```{r}
ds_dozen %>%
  filter(dataset=='away') %>%
  ggplot(aes(x=x, y =y, colour=dataset)) +
  geom_point()

```


##### Adjust the **filtering** step to plot two datasets

```{block, opts.label = "tip"}
R provides the inline instruction `%in%` to test if there a match of the left operand in the right one (a **vector** most probably)
```

```{r}
ds_dozen %>%
  filter(dataset %in% c('away', 'bullseye')) %>%
  ggplot(aes(x=x, y =y, colour=dataset)) +
  geom_point()
```


##### Expand now by getting one `dataset` per **facet**

Facet is applied in order to split the plots and separate datasets according to a variable.
```{r}
ds_dozen %>%
  filter(dataset %in% c('away', 'bullseye')) %>%
  ggplot(aes(x=x, y =y, colour=dataset)) +
  geom_point() +
  facet_wrap(~ dataset)

```


##### Remove the filtering step to facet all datasets

```{r, fig.height=8, fig.width=10}
ds_dozen %>%
  ggplot(aes(x=x, y =y, colour=dataset)) +
  geom_point() +
  facet_wrap(~ dataset)

```



##### Tweak the theme and use the `theme_void` and remove the legend

```{r, fig.height=8, fig.width=10}
ds_dozen %>%
  ggplot(aes(x=x, y =y)) +
  geom_point() +
  facet_wrap(~ dataset)+
  theme_void()

```


##### Are the datasets actually that similar?

No, the summary stats can be misleading


```{block, opts.label = "tip"}
the R package [`gifski`](https://cran.r-project.org/web/packages/gifski/index.html) could be installed on your machine, makes the GIF creation faster.
`gifski` is internally written in `rust`, and this language needs `cargo` to run. See this [article](https://github.com/r-rust/hellorust#installation) to get it installed on your machine.
First install `rust` before install the R package `gifski`.
Please note, that the `animate()` step still takes ~ 3-5 minutes depending on your machine.
```

##### Install `gganimate`, its dependencies will be automatically installed. {.bonus}

```{r}
# install.packages("gganimate")
# install.packages("rust")
# install.packages("gifski")
```


##### Use the `dataset` variable to the `transition_states()` argument layer {.bonus}

```{r}
library(gganimate)

ds_dozen %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  # transition will be made using the dataset column
  transition_states(dataset, transition_length = 5, state_length = 2) +
  # for a firework effect!
  shadow_wake(wake_length = 0.05) +
  labs(title = "dataset: {closest_state}") +
  theme_void(14) +
  theme(legend.position = "none") -> ds_anim
# more frames to slow down the animation
ds_gif <- animate(ds_anim, nframes = 500, fps = 10, renderer = gifski_renderer())
ds_gif
anim_save(title_frame = TRUE, "plots/ds.gif")
```



##### Visualize the tiny differences in means for both coordinates {.bonus}

- need to zoom tremendously to see differences. Accumulate all states to better see the motions.

```{r}
ds_dozen %>%
  group_by(dataset) %>%
  summarise(across(c(x, y), list(mean = mean, sd = sd))) %>% 
  ggplot(aes(x = x_mean, y = y_mean, colour = dataset)) +
  geom_point(size = 25, alpha = 0.6) +
  # zoom in like crazy
  coord_cartesian(xlim = c(54.25, 54.3), ylim = c(47.75, 47.9)) +
  # animate
  transition_states(dataset, transition_length = 5, state_length = 2) +
  # do not remove previous states to pile up dots
  shadow_mark() +
  labs(title = "dataset: {closest_state}") +
  theme_minimal(14) +
  theme(legend.position = "none") -> ds_mean_anim
ds_mean_gif <- animate(ds_mean_anim, nframes = 100, fps = 10, renderer = gifski_renderer())
ds_mean_gif
anim_save("plots/ds_mean.gif")
```



## Conclusion

> never trust summary statistics alone; always visualize your data | Alberto Cairo

**Authors**

- Alberto Cairo, (creator)
- Justin Matejka
- George Fitzmaurice
- Lucy McGowan

from this [post](https://itsalocke.com/datasaurus-now-cran/)
