---
title: "Wrapup Tidyverse"
author: "Elisa Gomez de Lope"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=FALSE}
library(tidyverse)
library(vroom)
library(here)
library(ggbeeswarm)
```




>This practical connects most lectures and practicals covered in the course as they would work together for a typical data analysis with data import, transformation, summarizing and plotting.

# Carbon dixoide plant intake

The dataset is from the paper by **Potvin, C., Lechowicz, M. J. and Tardif, S.** (1990). The statistical analysis of ecophysiological response curves obtained from experiments involving repeated measures., _Ecology_, **71**, 1389–1400. You can find it at this [URL](https://biostat2.uni.lu/practicals/data/CO2.tsv)

## Load the data

##### Load the [dataset](https://biostat2.uni.lu/practicals/data/CO2.tsv), assign the name `plant_CO2`

```{block, opts.label = "tip"}
Pay attention to use meaningful names. 
`plant_CO2_url` is a good name for the URL of this dataset.
`plant_CO2_file` is appropriate for tsv file if you download it.
```




```{r}
plant_CO2 <- vroom("data/CO2.tsv", col_types = cols())
```

## Explore the data

##### How many rows and columns are present? And which data types are they?

There are 84 rows, 5 columns
Column Plant, Type and Treatment are of type character;
Column conc, uptake are of type double


##### Three columns are categorical, taken together, how many observations have they?

```{block, opts.label = "tip"}
Using `count()` with the relevant columns is the easiest approach.
Sorting by the highest number is often a nice way to discover the data.
This option is available in `count()`
```

```{r}
count(plant_CO2)

plant_CO2 %>% 
   count(Plant, Type, Treatment, sort = T)

```


##### How many categories for each of the character columns do you find?

```{block, opts.label = "tip"}
One option is to find the number of unique values using `n_distinct()`.
This would have to be done three times, one for each character column. 
Otherwise, `n_distinct()` can be applied to character columns using
`across()` and relevant `where()` condition.
```

```{block, opts.label = "warning"}
`n_distinct()`, `across()`  are functions that work **inside** `mutate()` or `summarise()`.
```

```{r}
plant_CO2 %>%
   summarise(across(where(is.character), n_distinct))

```


##### Rename the columns to make them all lower case and continue with the resulting tibble for the next steps.

```{r}
plant_CO2 <- rename_with(plant_CO2,
            stringr::str_to_upper)


```




## CO<sub>2</sub> data

The numeric columns are defined as:

- `conc` is ambient carbon dioxide concentrations ($mL/L$).
- `uptake` is carbon dioxide uptake rates ($umol/m^2$ sec).

##### Compute the mean of `conc` and `uptake` depending on the location (Quebec or Mississippi)

```{block, opts.label = "tip"}
Ignore the treatment, chilled or / non chilled for now
```

```{r}
plant_CO2 %>%
   group_by(TYPE) %>%
   summarise(across(where(is.double),list(average = mean)))
```


##### Does the CO<sub>2</sub> concentrations appear different in both locations?

No, same concentrations in both locations


##### Compute the mean of `conc` and `uptake` depending on the plant location (`type` column), `plant` and treatment. Add the number of observations per group.

```{block, opts.label = "tip"}
the number of observations should be **7** everywhere as you saw before
```

```{r}
plant_CO2 %>%
   group_by(TYPE, PLANT, TREATMENT) %>%
   summarise(across(where(is.double), list(average = mean, n = n_distinct))) %>%
   select(-UPTAKE_n)

```


##### From the summary above, sort the table to show the plant / location / treatment that uptake the most carbon dioxide

```{block, opts.label = "tip"}
a descending ordering is possible with `desc(mean_uptake)` in the `arrange()` function
```


```{r}
plant_CO2 %>%
   group_by(TYPE, PLANT, TREATMENT) %>%
   summarise(across(where(is.double), list(average = mean, n = n_distinct))) %>%
   arrange(desc(UPTAKE_average)) %>%
   select(TYPE,PLANT,TREATMENT, UPTAKE_average)

```


It is hard to see if all plants have both treatment, you can pivot the table to find out. Right now the table after summarisation is in the **long** format. Pivoting means converting the table to the **wide format** using the treatment unique values (**chilled** and **nonchilled**) as new columns and filled with the values taken from the `mean_uptake` column.

##### Pivot the table using `plant` and `type` as ids, names from `treatment` and values from the mean of carbon dioxide uptake

```{block, opts.label = "tip"}
look up the different arguments of the function `tidyr::pivot_wider()`
```


```{r}
plant_CO2 %>%
   group_by(TYPE, PLANT, TREATMENT) %>%
   summarise(across(where(is.double), list(average = mean, n = n_distinct))) %>%
   arrange(desc(UPTAKE_average)) %>%
   select(TYPE,PLANT,TREATMENT, UPTAKE_average) %>%
   pivot_wider(names_from = TREATMENT, values_from = UPTAKE_average)
```


##### What do you conclude, does the same plant have been assessed for both treatment?

My conclusion is that the names in PLANT (i.e. Qn3) already contain the information of TYPE (Q ~ Quebec) and TREATMENT (n ~ nonchilled), besides an identification (3). So I think all plants (1, 2, 3) were assessed in both treatments (c,n) and in both locations (Q,M)


## Visualize the location and CO<sub>2</sub> uptake

Numbers are nice but plotting the data might better reveal how location and treatment relate.

##### Plot the CO<sub>2</sub> uptake in function of plant location, using violin plots filled by treatment {.bonus}

```{r}
plant_CO2 %>%
   select(TYPE, UPTAKE, TREATMENT) %>%
   ggplot(aes(x = TYPE,
              y = UPTAKE)) +
   geom_violin(aes(fill = TREATMENT), trim=F)
```


As a motivation, here is one version of the plot:



# Loading qPCR data from multiple files

- Download a set of qPCR experiments results [here as a zip archive](https://biostat2.uni.lu/practicals/data/pcr.zip) to the `data` sub-folder.
- Decompress the zip archive to the `data` sub-folder of your project.

```{block, title = "test", opts.label = "tip"}
You can run the following command in R to decompress the file:

~~~r
unzip(here::here("data", "pcr.zip"), exdir = "data/pcr")
~~~

This will create a sub-folder `data/pcr` and extract the files there.
```


These qPCR results were obtained from 2 different samples and were replicated 5 times each. They are stored in **10** different files. A filename looks like `mlc1_1.csv` where

- the first number is the **sample id**
- the second the **replicate id**.

#### Check that your file locations are correct

```{r}
dir(here::here("data", "pcr"), full.names = FALSE)
```


## Read in the qPCR results


The files exhibits a `csv` extension: you might think that the delimiter is a comma.

Let `vroom` doing the job for us.

##### Use `vroom` to import the file `mlc1_1.csv`

```{r}
vroom("data/pcr/mlc1_1.csv")
```

The delimiter was a space (" ")


## Get a single tibble tidy out of all files

Instead of getting a list of multiple tibbles we can bind them altogether to create a single one.

`vroom` can actually achieve this if the input is a list of file paths, it merges them into **one** `tibble` **and** allows for keeping track of the **filename**.

An optional `id` option allows to save the `path` to keep track of the file of origin. Let's name this `id` column `"file"` and save the resulting tibble of 96x10 rows as `pcr_data`

##### Use `dir()` to create a vector named `pcr_files` which contains the path to all 10 data files

```{block, opts.label = "tip"}
You can get some help using `?dir` in particular to adjust the `full.names` argument to `TRUE`.

Then, we obtain the full path to each file, necessary to access the data.
```

```{r}
vroom(dir(path = "data/pcr", full.names = T), id = "file")
#vroom(fs::dir_ls("data/pcr/", full.names = T), id = "file") also worked

```


##### Read all 10 files at once with `vroom`, using `id = "file"` to add filenames as an extra column. Assign the name `pcr_data`

```{r}
pcr_data <- vroom(dir(path = "data/pcr", full.names = T), id = "file")

```


##### Is this dataset tidy?

No, samples and replicates should have their own column


##### Extract the sample (`mlc1` or `mlc2`) and replicate number (1 to 5) into dedicated columns, assign the name `pcr_data_tidy`

```{block, opts.label = "tip"}
Several options to achieve this are available. The most elegant is `str_extract()` with appropriate regular expressions. You can also get rid of the `file` column afterwards.

But extracting the filenames without extension and use `separate()` is also an alternative. 

For the regular expression, if you want to extract the digit **after** the `_` you can use the positive look-behind assertion: `"(?<=_)\\d"`
```


```{r}
pcr_data_tidy <- pcr_data %>%
   mutate(sample = str_extract(file, ".{3}[0-9]{1}"),
          replicate = str_extract(file, "(?<=_)\\d+")) %>%
   select(-file)
```



Of note, the resulting **tidy** `tibble` should be:

```
# A tibble: 960 x 7
   well  gene      Ct type  flag  sample replicate
   <chr> <chr>  <dbl> <chr> <chr> <chr>  <chr>    
 1 A1    gene1   33.6 data  OK    mlc1   1        
 2 A2    gene2   22.4 data  OK    mlc1   1        
 3 A3    gene3   31.8 data  OK    mlc1   1        
 4 A4    gene4   30.2 data  OK    mlc1   1        
 5 A5    gene5   33.1 data  OK    mlc1   1        
 6 A6    gene6   33.8 data  OK    mlc1   1        
 7 A7    gene7   31.0 data  OK    mlc1   1        
 8 A8    gene8   16.3 data  OK    mlc1   1        
 9 A9    gene9   32.6 data  OK    mlc1   1        
10 A10   gene10  30.1 data  OK    mlc1   1        
# … with 950 more rows
```

##### Check that for both samples, all `flag` are **OK** by counting them

```{r}
pcr_data_tidy %>%
   group_by(sample) %>%
   count(flag)
```


##### Now that the data is tidy, plot the distribution of Ct values

+ `replicate` on the x-axis
+ `sample` as facet (2 columns)
+ geometries, you can combine both a violin plot and the `ggbeeswarm::geom_quasirandom`

```{r}
pcr_data_tidy %>%
   ggplot(aes(x = replicate,
              y = Ct)) +
   geom_violin(trim = F) +
   facet_wrap(~ sample) +
   ggbeeswarm::geom_quasirandom(aes(alpha = 0.6))

```


