---
title: "Tidyverse import data"
author: "Elisa Gomez de Lope"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---


```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=FALSE}
library(dplyr)
library(vroom)
```



> In this practical, you'll learn how to import flat files using the `vroom` package


##### Those kind of questions are optional {.bonus}

# Before you start


To perform reproducible research it is a good practice to store the files in a standardized location. For example, you could take advantage of the RStudio projects and store data files in a sub-folder called `data`.

If you did not create an Rstudio project yet, **create a RStudio project**.

##### Prepare your project's folder

1. Check that the project is active: the name you chose should appear on the top-right corner.
    
2. Create a folder named `data` within your project's folder. Use the `Files` pane in the lower right Rstudio panel or your favorite file browser.

3. Download the file [`blood_fat.csv`](https://biostat2.uni.lu/practicals/data/blood_fat.csv) and place it in the `data` sub-folder you just created.

```{block show_wd, opts.label = "advice"}
Use `here::here()` to determine your project's root folder.\
Go to the path showed by one of these commands to find your files.
```

4. Create a new **Rmarkdown** file, save it at the project root with a relevant name.

5. Edit the first code chunk to include
library(dplyr)
library(vroom)

6. Don't forget to run the chunk's code to load the library during your interactive session

```{block, opts.label = "warning"}
If you load the library only in the console and forget to place a chunk to load it, the knitting process will fail. Indeed, when you click on the `knit` button, the chunks are evaluated in a new and fresh environment.
```

# Use `vroom` to load your first file


##### Load the `blood_fat` file

```{block, opts.label = "tip"}
the relative path can be safely built using `here::here("data", "blood_fat.csv")` 
if you followed the preliminary steps above, download the csv in a sub-folder `data` of a RStudio project

For example, you folder structure could be (depending on the picked names). Here:

- RStudio project is `rworkshop`
- Rmarkdown document is `practical02_import.Rmd`

~~~
.
├── data
│   └── blood_fat.csv
├── practical03_import.Rmd
└── rworkshop.Rproj
~~~

```


```{r}
vroom("data/blood_fat.csv")
```



`vroom` is reporting the dimensions of the file, along with the guessed delimiter and data type of each columns

If we are happy with the guessed delimiter and the column names / types, we could silent this reporting.

##### Load again the same file, silencing the `vroom` message

```{r}
blood_fat <- vroom("data/blood_fat.csv")

```


## The tibble

`vroom()` loads the data as a tibble. The main advantage to use tibbles over a regular data frame is the printing.

- Tibbles show some useful information such as the number of rows and columns:
    + Look at the top of the tibble and find the information "A tibble rows x cols"
    + How many rows are in the tibble?
- The columns of a tibble report their type:
    + Look at the tibble header, the type of a columns is reported just below its name.
    + What is the type of the `age` column?
    
There are 25 rows, the type of age column is double.


Actually, both `age` and `id` are integers, and should be read as such.

##### Read the `blood_fat.csv` specifying the data types of `age` and `id` as integers


```{block, opts.label = "tip"}
In the `col_Types = cols()` you can use the columns bare names and either the long description
 to call the specific data type like `col_integer()` or the shortcut `"i"`
```

```{r}
blood_fat <- vroom("data/blood_fat.csv",
                   col_types = cols(age= "i", id="i"))
```


##### Read the `blood_fat.csv` specifying the data types of `age` and `id` as integers, skipping `weight`

```{r}
blood_fat <- vroom("data/blood_fat.csv",
                   col_types = cols(age= "i", id="i", weight= "_"))
```


##### One summarisation: compute the **mean** of both the `age` and `fat` per `group` {.bonus}

```{r}
blood_fat %>%
  group_by(group) %>%
  summarise(across(c(age, fat), list(mean = mean, sd = sd)))
```

